Automatic Identification and Annotation of Parallel Tasks in Structured Programs

### 2017-12-06 docs/ACM/paper.tex
Working a bit further on Section 2.

The main advantage of using OpenMP to split a program into parallel tasks is
its runtime.
This runtime lets us circumvent the shortcomings that pointer aliasing
impose on the automatic parallelization of code.
Aliasing -- the possibility of two pointers dereferencing the same
memory region -- either prevents parallelism altogether, or forces compilers to
resort to complex runtime checks to identify it
dynamically~\cite{Alves15,Rus02}.
The OpenMP runtime shields us from having to implement such checks, because
it already checks for dependencies among tasks, and dispatches them in a
correct order~\cite{LaGrone11}.
An OpenMP task descriptor contains the task's status (running, waiting) and a
list of memory addresses the task registered as input arguments.

The process of checking for dependencies can be as simple as checking if an
incoming task in the system accesses a memory address already in use by a
previous task in flight.
It can also be complex, involving checks over long ranges of addresses
(in case the runtime system supports dependence checking for memory ranges).
This process can also include memory labelling and renaming (in case the
runtime system is able to remove false dependencies).
Regardless of its capacity -- which is not the same across every OpenMP
implementation -- the runtime system will represent dependencies using a task
dependence graph (TDG)~\cite{Duran08}.
In this directed acyclic graph, nodes denote tasks and edges represent
dependences between them.
Tasks are dispatched for execution according to a topological ordering of this
graph~\cite{Planas15}.

This runtime support allows the parallelization of irregular applications,
such as programs that traverse data-structures formed by a mesh of pointers.
As an example, Figure~\ref{fig:book_filter} shows an application that
finds patterns in lines of a book.
The book is given as an array of pointers; each pointer leads to a
string representing a --potentially-- different line.
The natural parallelization of this program consists in firing off a task
to process each line.
Figure~\ref{fig:book_filter} has been annotated automatically by the \Taskminer.
We emphasize that neither the runtime checks of Alves
{\em et al.}~\cite{Alves15} or Rus {\em et al.}~\cite{Rus02}, nor a
context and flow sensitive alias analysis, \`{a} la Whaley and
Lam~\cite{Whaley04} would be able to ensure the automatic parallelization of
this program.

@@

The creation of tasks involves a non-negligible runtime cost due to allocation,
scheduling of resources and real-time management of the task dependence graph.
Ideally, this cost should be paid only for tasks that perform an amount of work
sufficiently large to pay for their management.
Being an interesting program property, on Rice's sense~\cite{Rice53}, the amount
of work performed by a task cannot be discovered statically.
As we show in Section~\ref{sub:profit}, we can try to approximate this quantity, using,
to this end, program symbols, which are replaced with actual values at runtime.
For instance, in Figure~\ref{fig:ex_Regions}, we know that the body of the
innermost loop is formed by five instructions.
Thus, we approximate the amount of work performed by a task with the
expression \textsf{5 * M}.
We use the runtime value of \textsf{M} to determine, during the execution of the
program, if we create a task, or not.
Such test is carried out by the guard at line 7 of the figure, which is part of
OpenMP's syntax. Also, we provide a reliable estimate on the \emph{workload cutoff}
from which a task can be safely spawned without producing performance overhead.
This \emph{cutoff} considers factors such as number of available cores
and runtime information on the task dispatch cost in terms of machine
instructions. This is further explained in Section \ref{sub:profit}.

\begin{figure}[h!]
\begin{center}
\includegraphics[width=0.95\columnwidth]{images/ex_cutoff}
\caption{Challenge~\ref{ch:cutoff}: bounding the creation of recursive tasks with
counter associated with function calls.}
\label{fig:ex_cutoff}
\end{center}
\end{figure}

>>>>

The main advantage of using OpenMP to split a program into parallel tasks is
its runtime.
This runtime lets us circumvent the shortcomings that pointer aliasing
impose on the automatic parallelization of code.
Aliasing -- the possibility of two pointers dereferencing the same
memory region -- either prevents parallelism altogether, or forces compilers to
resort to complex runtime checks to ensure its correctness~\cite{Alves15,Rus02}.
The OpenMP runtime shields us from this problem, because
it already checks for dependencies among tasks, and dispatches them in a
correct order~\cite{LaGrone11}.
An OpenMP task descriptor contains the task's status (running, waiting) and a
list of memory addresses the task registered as input arguments.

The process of checking for dependencies can be as simple as checking if an
incoming task accesses a memory address already in use by a task in flight.
It can also be complex, involving checks over long ranges of addresses
(in case the runtime system supports dependence checking for memory ranges).
This process can also include memory labelling and renaming (in case the
runtime system is able to remove false dependencies).
Regardless of its capacity -- which is not the same across every OpenMP
implementation -- the runtime system will represent dependencies using a task
dependence graph (TDG)~\cite{Duran08}.
In this directed acyclic graph, nodes denote tasks and edges represent
dependences between them.
Tasks are dispatched for execution according to a topological ordering of this
graph~\cite{Planas15}.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=0.9\columnwidth]{images/ex_book_filter}
\caption{The benefits of the OpenMP runtime environment.
Neither the runtime checks of Alves~\cite{Alves15} or Rus~\cite{Rus02}, nor
Whaley's context and flow sensitive alias analysis would be able to ensure the
correctness of the automatic parallelization of this program.}
\label{fig:ex_book_filter}
\end{center}
\end{figure}

This runtime support allows the parallelization of irregular applications,
such as programs that traverse data-structures formed by a mesh of pointers.
As an example, Figure~\ref{fig:ex_book_filter} shows an application that
finds patterns in lines of a book.
The book is given as an array of pointers; each pointer leads to a
string representing a --potentially-- different line.
The natural parallelization of this program consists in firing off a task
to process each line.
Figure~\ref{fig:ex_book_filter} has been annotated automatically by the
\Taskminer.
The OpenMP execution environment ensures the correct scheduling of the tasks
created at lines 9 and 11, through runtime checks.
Therefore, the ability to disambiguate memory regions statically is not
essential for correctness, although it is important for efficiency reasons,
as we will mention in Challenge~\ref{ch:Regions}.

@@

The creation of tasks involves a heavy runtime cost due to allocation,
scheduling and real-time management of the task dependence graph.
Ideally, this cost should be paid only for tasks that perform an amount of work
sufficiently large to pay for their management.
Being an interesting program property, on Rice's sense~\cite{Rice53}, the amount
of work performed by a task cannot be discovered statically.
As we show in Section~\ref{sub:profit}, we can try to approximate this quantity, using,
to this end, program symbols, which are replaced with actual values at runtime.
For instance, in Figure~\ref{fig:ex_Regions}, we know that the body of the
innermost loop is formed by five instructions.
Thus, we approximate the amount of work performed by a task with the
expression \textsf{5 * M}.
We use the runtime value of \textsf{M} to determine, during the execution of the
program, if we create a task, or not.
Such test is carried out by the guard at line 7 of the figure, which is part of
OpenMP's syntax. Also, we provide a reliable estimate on the \emph{workload cutoff}
from which a task can be safely spawned without producing performance overhead.
This \emph{cutoff} considers factors such as number of available cores
and runtime information on the task dispatch cost in terms of machine
instructions. This is further explained in Section \ref{sub:profit}.

\begin{figure}[h!]
\begin{center}
\includegraphics[width=0.9\columnwidth]{images/ex_cutoff}
\caption{Challenge~\ref{ch:cutoff}: bounding the creation of recursive tasks with
counter associated with function calls.}
\label{fig:ex_cutoff}
\end{center}
\end{figure}
